Recursive Solutions to First-Order Model Counting

### 2022-08-01 doc/paper/paper.tex
Clean up packages & commands
Improve wording

\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{subcaption}
\usepackage[inline,shortlabels]{enumitem}
\usepackage{complexity}

@@

\usetikzlibrary{fit,positioning,trees,cd}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\newcommand{\crefrangeconjunction}{--}
\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\IE}{\textsc{IE}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\wwp}{w}
\DeclareMathOperator{\wwn}{\overline{w}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\size}{\sigma}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\gr}{gr}
\DeclareMathOperator{\leftlsquigarrow}{\text{\reflectbox{$\rightsquigarrow$}}}
\SetKw{KwRet}{yield}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{applyAllRules}{applyAllRules}
\SetKwFunction{applyGreedyRules}{applyGreedyRules}
\SetKwFunction{applyGreedyRulesToFormulas}{applyGreedyRulesToFormulas}
\SetKwFunction{emptyy}{empty}
\SetKwFunction{get}{get}
\SetKwFunction{put}{put}
\SetKwFunction{constructDomainMap}{constructDomainMap}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}
\SetKwFunction{mergeFcgs}{mergeFcgs}
\SetKwFunction{updateCache}{updCache}
\newtheorem{assumption}{Assumption}
\newtheorem{conjecture}{Conjecture}
\newtheorem{constraint}{Constraint}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{experiment}{Experiment}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\Crefname{constraint}{Constraint}{Constraints}
\Crefname{experiment}{Experiment}{Experiments}
\Crefname{clause}{Clause}{Clauses}
\Crefname{diagram}{Diagram}{Diagrams}
\crefname{line}{line}{lines}

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{The problem of computing
  the factorial can be described using two variables and counting quantifiers,
  so it is known to be liftable in principle
  \citep{DBLP:journals/jair/Kuzelka21}.} We claim that this shortcoming is due
to the inability of these algorithms to construct recursive solutions. The topic
of recursion in the context of WFOMC has been studied before but in limited
ways. \Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data
which is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift} written in Scala. \textsc{ForcLift}
works in two stages: compilation and evaluation/propagation.\footnote{There is
  also an intermediate stage called \emph{smoothing} that takes place between
  compilation and evaluation. As our changes to smoothing are quite elementary,
  we do not discuss them to not distract the reader from the main contributions
  of this paper.} In the first part, various \emph{(compilation) rules} are
applied to the input (or some derivative) formula, gradually constructing a
circuit. In the second part, the weights of the instance (and sometimes
constants) are propagated through the circuit, computing the WMC\@.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} is able to translate
sentences in a variant of function-free many-sorted FOL with equality to this
internal format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$
and terms $t_{1}, \dots, t_{n}$. Here, $n \in \mathbb{N}_0$ is the \emph{arity}
of $p$. A \emph{term} is either a constant or a variable. A \emph{literal} is
either an atom or the negation of an atom (denoted by
$\neg p(t_1, \dots, t_n)$). Let $\mathcal{D}$ be the set of all domains.
Initially, $\mathcal{D}$ contains all domains mentioned by the input (W)FOMC
instance. During compilation, new domains added to $\mathcal{D}$ by some of the
compilation rules.

@@

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $p$, variables $X$, $Y$, $Z$, and domains $\Gamma$ and
  $\Delta$. Then $\Vars(\{\, (Y, Z) \,\}) = \{\, Y, Z \,\}$, and
  $\Vars(c_{1}) = \Vars(c_{2}) = \{\, X, Y, Z \,\}$. Based on the domain maps of
  $c_{1}$ and $c_{2}$, we can infer that $p \in \Gamma \times \Delta$. All
  variables (in both clauses) that occur as the first argument to $p$ are in
  $\Gamma$, and likewise all variables that occur as the second argument to $p$
  are in $\Delta$. Therefore, $\phi$ `type checks' as a valid formula.
\end{example}

@@

One can read a formula as in \cref{def:formula} as a sentence in a FOL\@. All
variables in a clause are implicitly universally quantified (but note that
variables are never shared among clauses), and all clauses in a formula are
implicitly linked by conjunction. Thus, formula $\phi$ from \cref{example:first}
reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ to denote the length of list
$l$. We write $\pfun$ for partial functions. Let $S$ be a set of constraints or
literals, $V$ a set of variables, and $x$ either a variable or a constant. We
write $S[x/V]$ to denote $S$ with all occurrences of all variables in $V$
replaced with $x$.\footnote{Note that if $(X, Y)$ is a two-variable constraint,
  substituting a constant $c$ for $X$ would result in $(c, Y)$, which would have
  to be rewritten as $(Y, c)$ to fit the definition of a constraint.}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit nodes whose only arithmetic effect is multiplication by
    one. Some of these nodes play an important role in the weighted version of
    the problem whereas others are remnants of the interaction between
    compilation rules and the way in which \textsc{ForcLift} handles existential
    quantifiers.}\label{fig:examplefcg}
\end{figure}

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges.\footnote{Note that
  imposing an ordering on outgoing edges is just a limited version of edge
  labelling.} Node labels consist of two parts: the \emph{type} and the
\emph{parameters}. The out-degree of each node is determined by its type.
Similarly to \citet{DBLP:conf/ijcai/BroeckTMDR11}, we write $T_p$ for an FCG
that has a node with label $T_p$ (i.e., type $T$ and parameter(s) $p$) and
$\star$'s as all of its direct successors.

Most of the types are as in previous work
\citep{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}. The type for
non-tree-like edges (denoted by $\Reff$), while used before, is extended to
contain the information necessary to support recursive calls. We also add three
new types:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
When drawing an FCG, we order outgoing edges from left to right, write node
labels directly on the nodes, and omit irrelevant labels and/or parameters. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1,
label $\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

% Connecting FCGs with formulas
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, s.t. $|L|$ is equal to the number
of $\star$'s in $G$. The intuition behind this definition is that $L$ contains
formulas that still need to be compiled. Once a formula is compiled, it replaced
one of the $\star$'s in $G$ according to a set order. We say that an FCG is
\emph{complete} (i.e., it represents a \emph{complete solution}) if it has no
$\star$'s. Similarly, a chip is complete if its FCG is complete (or,
equivalently, the list of formulas is empty).

@@

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While \textsc{ForcLift}
\citep{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one of them, in an
attempt to not miss a solution, \textsc{Crane} returns them all. In particular,
if a rule returns an empty set, then that rule is not applicable to the formula,
and the algorithm continues with the next rule.

@@

  \Cref{line:conditions} iterates over all subsets $W \subseteq V$ of variables
  that can be replaced by a constant without resulting in formulas that are
  evidently unsatisfiable. We impose two restrictions on $W$. First,
  $W^2 \cap C = \emptyset$ ensures that there are no pairs of variables in $W$
  that are constrained to be distinct, since that would result in a $x \ne x$
  constraint after substitution. Similarly, we want to avoid variables in $W$
  that have inequality constraints with constants: after substitution, such
  constraints would transform into inequality constraints between two constants.
  In this case, both subsets of $V$ satisfy these conditions, and
  \cref{line:generation} generates two clauses for the output formula:
  \begin{multline*}
    (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z), (X, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
  \end{multline*}
  from $W = \emptyset$ and
  \[
    (\{\, \neg p(x, Y), \neg p(x, Z) \,\}, \{\, (Y, Z) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\})
  \]
  from $W = V$.

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2, c_3 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &= (\emptyset, \{\, (Y, X) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\}), \\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_3 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and with its element $x \in \Gamma$ satisfy the preconditions
  for CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2', c_3')$, where
  \begin{align*}
    c_1' &= c_1, \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_3' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

\citet{DBLP:conf/ijcai/BroeckTMDR11} use a cache to check if a formula is
identical to one of the formulas that have already been fully compiled. In
\textsc{Crane}, to facilitate recursion, we extend the caching scheme to include
formulas that have been encountered but not fully compiled yet. Formally, we
define a \emph{cache} to be a map from integers (e.g., hash codes) to sets of
pairs of the form $(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG
node.

@@

% the process: both for running algorithms and for determining complexities
In our experiments, we consider all sixteen combinations of these properties,
i.e., injectivity, surjectivity, partiality, and endo-. \textsc{ForcLift} is
always run until it terminates. \textsc{Crane} is run until either five
solutions are found or the search tree reaches height 6.\footnote{The search
  tree has a high branching factor, so exploring all nodes at depth 5 takes at
  most a few seconds whereas doing the same for depth 6 can be computationally
  infeasible in some cases.} If successful, \textsc{ForcLift} generates a
circuit, and \textsc{Crane} generates one or more (complete) FCGs. In both
cases, we manually convert the resulting graphs into definitions of functions as
described in \cref{sec:interpret}. We then assess the complexity of each
solution and pick the best if \textsc{Crane} returns several solutions of
varying complexities. When assessing the complexity of each such definition, we
make two assumptions. First, we can compute the binomial coefficient
$\binom{n}{k}$ in $\Theta(nk)$ time. Second, techniques such as dynamic
programming and memoization are used to avoid recomputing the same binomial
coefficient or function call multiple times.

>>>>

\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}

@@

\usetikzlibrary{cd}
\usetikzlibrary{shapes}

\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{identifyRecursion}{r}
\SetKwFunction{generateMaps}{genMaps}
\newtheorem{conjecture}{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{line}{line}{lines}

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{The problem of computing
  the factorial can be described using two variables and counting quantifiers,
  so it is known to be liftable in principle
  \citep{DBLP:journals/jair/Kuzelka21}.} We claim that this shortcoming is due
to the inability of these algorithms to construct recursive solutions. The topic
of recursion in the context of WFOMC has been studied before but in limited
ways. \Citet{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical data
that is then used to conjecture recurrence relations that explain that data.
\Citet{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain recursion}.
Intuitively, domain recursion partitions a domain of size $n$ into a single
explicitly named constant and the remaining domain of size $n-1$. However, many
stringent conditions are enforced to ensure that the search for a tractable
solution always terminates.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift} written in Scala. \textsc{ForcLift}
works in two stages: compilation and evaluation/propagation.\footnote{There is
  also an intermediate stage called \emph{smoothing} that takes place between
  compilation and evaluation. As our changes to smoothing are rather elementary,
  we do not discuss them to not distract the reader from the main contributions
  of this paper.} In the first part, various \emph{(compilation) rules} are
applied to the input (or some derivative) formula, gradually constructing a
circuit. In the second part, the weights of the instance (and sometimes
constants) are propagated through the circuit, computing the WMC\@.

@@

Our representation of FOMC instances is largely based on the format used
internally by \textsc{ForcLift}, some aspects of which are described by
\citet{DBLP:conf/ijcai/BroeckTMDR11}. \textsc{ForcLift} can translate sentences
in a variant of function-free many-sorted FOL with equality to this internal
format. An \emph{atom} is $p(t_1, \dots, t_n)$ for some predicate $p$ and terms
$t_{1}, \dots, t_{n}$. Here, $n \in \mathbb{N}_0$ is the \emph{arity} of $p$. A
\emph{term} is either a constant or a variable. A \emph{literal} is either an
atom or the negation of an atom (denoted by $\neg p(t_1, \dots, t_n)$). Let
$\mathcal{D}$ be the set of all domains. Initially, $\mathcal{D}$ contains all
domains mentioned by the input (W)FOMC instance. During compilation, new domains
are added to $\mathcal{D}$ by some of the compilation rules.

@@

\begin{example}\label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
          \end{multlined}\\
    c_2 &\coloneqq
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Gamma \,\})
          \end{multlined}
  \end{align*}
  for some predicate $p$, variables $X$, $Y$, $Z$, and domains $\Gamma$ and
  $\Delta$. Then $\Vars(\{\, (Y, Z) \,\}) = \{\, Y, Z \,\}$, and
  $\Vars(c_{1}) = \Vars(c_{2}) = \{\, X, Y, Z \,\}$. Based on the domain maps of
  $c_{1}$ and $c_{2}$, we can infer that $p \in \Gamma \times \Delta$. All
  variables (in both clauses) that occur as the first argument to $p$ are in
  $\Gamma$, and, likewise, all variables that occur as the second argument to
  $p$ are in $\Delta$. Therefore, $\phi$ `type checks' as a valid formula.
\end{example}

@@

One can read a formula from \cref{def:formula} as a sentence in a FOL\@. All
variables in a clause are implicitly universally quantified (but note that
variables are never shared among clauses), and all clauses in a formula are
implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \implies \neg p(X, Y) \lor \neg p(X, Z)) \land
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \implies \neg p(X, Y) \lor \neg p(Z, Y)).
    \end{multlined}
\end{align*}

@@

\paragraph*{Notation.}
We write $\langle\rangle$ and $\langle x \rangle$ to denote an empty list and a
list with one element $x$, respectively, and $|l|$ for the length of list $l$.
We write $\pfun$ for partial functions. Let $S$ be a set of constraints or
literals, $V$ a set of variables, and $x$ a variable or a constant. We write
$S[x/V]$ to denote $S$ with all occurrences of all variables in $V$ replaced
with $x$.\footnote{Note that if $(X, Y)$ is a two-variable constraint,
  substituting a constant $c$ for $X$ would result in $(c, Y)$, which would have
  to be rewritten as $(Y, c)$ to fit the definition of a constraint.}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[every node/.style={draw,ellipse},edge from
    parent/.style={draw,-Latex},sibling distance=3cm]
    \node (dr) {$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$}
    child {node {$\bigvee_{\Delta^\top \subseteq \Delta}$}
      child {node {$\CR_{\Gamma \mapsto \Gamma'}$}
        child {
          node[circle] {$\wedge$}
          child {node {$\bot$}}
          child {node (ref) {$\Reff_{\{\, \Gamma \mapsto \Gamma', \Delta \mapsto \Delta^{\bot} \,\}}$}}
        }}};
    \draw[-Latex, bend right] (ref) to (dr);
  \end{tikzpicture}
  \caption{A simplified version of an FCG constructed by \textsc{Crane} for the
    problem of counting partial injections from \cref{example:first}. Label
    $\bigvee_{\Delta^\top \subseteq \Delta}$ denotes set-disjunction, $\land$
    denotes conjunction, and $\bot$ denotes a contradiction---see the work by
    \citet{DBLP:conf/ijcai/BroeckTMDR11} for the descriptions of these node
    types. Here we omit nodes whose only arithmetic effect is multiplication by
    one. Some of these nodes play an important role in the weighted version of
    the problem, whereas others are remnants of the interaction between
    compilation rules and the way in which \textsc{ForcLift} handles existential
    quantifiers.}\label{fig:examplefcg}
\end{figure}

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges.\footnote{Note that
  imposing an ordering on outgoing edges is just a limited version of edge
  labelling.} Node labels consist of two parts: the \emph{type} and the
\emph{parameters}. The type of a node determines its out-degree. Similarly to
\citet{DBLP:conf/ijcai/BroeckTMDR11}, we write $T_p$ for an FCG that has a node
with the label $T_p$ (i.e., type $T$ and parameter(s) $p$) and $\star$'s as all
of its direct successors.

Most node types are as in previous work
\citep{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}. The type for
non-tree-like edges (denoted by $\Reff$), while used before, is extended to
contain the information necessary to support recursive calls. We also add three
new types:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
When drawing an FCG, we order outgoing edges from left to right, write node
labels directly on the nodes, and omit irrelevant labels and/or parameters. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1,
label $\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

% Connecting FCGs with formulas
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, s.t. $|L|$ is equal to the number
of $\star$'s in $G$. The intuition behind this definition is that $L$ contains
formulas that still need to be compiled. Once a formula is compiled, it replaces
one of the $\star$'s in $G$ according to a set order. We say that an FCG is
\emph{complete} (i.e., it represents a \emph{complete solution}) if it has no
$\star$'s. Similarly, a chip is complete if its FCG is complete (or,
equivalently, the list of formulas is empty).

@@

A \emph{(compilation) rule} takes a formula and returns a set of chips. The
cardinality of this set is the number of different ways in which the rule can be
applied to the input formula. While \textsc{ForcLift}
\citep{DBLP:conf/ijcai/BroeckTMDR11} heuristically chooses one of them, in an
attempt to not miss a solution, \textsc{Crane} returns them all. In particular,
if a rule returns an empty set, then that rule does not apply to the formula,
and the algorithm continues with the next rule.

@@

  \Cref{line:conditions} iterates over all subsets $W \subseteq V$ of variables
  that can be replaced by a constant without resulting in evidently
  unsatisfiable formulas. We impose two restrictions on $W$. First,
  $W^2 \cap C = \emptyset$ ensures that there are no pairs of variables in $W$
  that are constrained to be distinct, since that would result in an $x \ne x$
  constraint after substitution. Similarly, we want to avoid variables in $W$
  that have inequality constraints with constants: after the substitution, such
  constraints would transform into inequality constraints between two constants.
  In this case, both subsets of $V$ satisfy these conditions, and
  \cref{line:generation} generates two clauses for the output formula:
  \begin{multline*}
    (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z), (X, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
  \end{multline*}
  from $W = \emptyset$ and
  \[
    (\{\, \neg p(x, Y), \neg p(x, Z) \,\}, \{\, (Y, Z) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\})
  \]
  from $W = V$.

@@

\begin{example}
  Let $\phi = \{\, c_1, c_2, c_3 \,\}$ be a formula with clauses
  \begin{align*}
    c_1 &= (\emptyset, \{\, (Y, X) \,\}, \{\, X \mapsto \Delta^\top, Y \mapsto \Delta^\top \,\}), \\
    c_2 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (X, x), (Y, Z) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
          \end{multlined}\\
    c_3 &=
          \begin{multlined}[t]
            (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \\
            \{\, (X, x), (Z, X), (Z, x) \,\}, \\
            \{\, X \mapsto \Gamma, Y \mapsto \Delta^\bot, Z \mapsto \Gamma \,\}).
          \end{multlined}
  \end{align*}
  Domain $\Gamma$ and its element $x \in \Gamma$ satisfy the preconditions for
  CR\@. The rule introduces a new domain $\Gamma'$ and transforms $\phi$ to
  $\phi' = (c_1', c_2', c_3')$, where
  \begin{align*}
    c_1' &= c_1, \\
    c_2' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Delta^\bot \,\}),
           \end{multlined} \\
    c_3' &=
           \begin{multlined}[t]
             (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \\
             \{\, X \mapsto \Gamma', Y \mapsto \Delta^\bot, Z \mapsto \Gamma' \,\}).
           \end{multlined}
  \end{align*}
\end{example}

@@

\citet{DBLP:conf/ijcai/BroeckTMDR11} use a cache to check if a formula is
identical to one of the formulas that have already been fully compiled. To
facilitate recursion, we extend the caching scheme to include formulas that have
been encountered but not fully compiled yet. Formally, we define a \emph{cache}
to be a map from integers (e.g., hash codes) to sets of pairs of the form
$(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG node.

@@

% the process: both for running algorithms and for determining complexities
In our experiments, we consider all sixteen combinations of these properties,
i.e., injectivity, surjectivity, partiality, and endo-. \textsc{ForcLift} is
always run until it terminates. \textsc{Crane} is run until either five
solutions are found or the search tree reaches height 6.\footnote{The search
  tree has a high branching factor, so exploring all nodes at depth 5 takes at
  most a few seconds, whereas doing the same for depth 6 can be computationally
  infeasible in some cases.} If successful, \textsc{ForcLift} generates a
circuit, and \textsc{Crane} generates one or more (complete) FCGs. In both
cases, we manually convert the resulting graphs into definitions of functions as
described in \cref{sec:interpret}. We then assess the complexity of each
solution and pick the best if \textsc{Crane} returns several solutions of
varying complexities. When assessing the complexity of each such definition, we
make two assumptions. First, we can compute the binomial coefficient
$\binom{n}{k}$ in $\Theta(nk)$ time. Second, techniques such as dynamic
programming and memoization are used to avoid recomputing the same binomial
coefficient or function call multiple times.
